Which groups are at highest risk for low uptake of flu vaccine?

### 2023-04-21 products/manuscript/Manuscript.qmd

---
title: "Which groups are at highest risk for low uptake of flu vaccine?"
author: Chris Okitondo and Irene Cavros
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

Influenza, or flu, is an extremely contagious respiratory illness caused
by a virus. It can cause a range of symptoms, including fever, cough,
sore throat, body aches, headache, chills, and fatigue (CDC, Key Facts
about Seasonal Flu Vaccine, 2022). It can be particularly dangerous for
certain groups, such as older adults, young children, and individuals
with underlying health issues, often resulting in severe illness or even
death  Although influenza vaccinations are widely accessible, many
people in the United States do not receive them as recommended. This
study examines influenza vaccination uptake among US adults from
2012-2021. Using a subset of the National Health Interview Survey (NHIS)
data set, researchers provide important insights into which groups are
at highest risk for under-coverage of flu vaccine in order to better
inform targeted public health interventions for improving uptake.

@@

### General Background Information

Influenza, commonly known as the flu, is a contagious respiratory
illness which comes from influenza viruses. It can cause mild to severe
illness, and in some cases, it can lead to hospitalization or even
death. Symptoms of the flu include fever, cough, sore throat, body
aches, headache, chills, and fatigue (CDC, About Flu, 2022). It can
spread easily from person to person through droplets that are released
when infected individuals cough, sneeze, or talk.

Vaccines are a crucial tool in preventing the spread of infectious
diseases, including the flu. On an individual level, the best way to
protect against the flu is by getting vaccinated each year. At the
population level, high vaccine coverage is a powerful strategy for
reducing spread of the virus (CDC, Key Facts about Seasonal Flu Vaccine,
2022).

The U.S. Centers for Disease Control and Prevention (CDC) recommends
that everyone six months of age and older receive a flu vaccine each
year, with rare exceptions (CDC, Key Facts about Seasonal Flu Vaccine,
2022). Influenza is a constantly evolving virus that mutates rapidly,
and uncontrolled spread gives rise to many different strains, so new
vaccines are developed each year to protect against the most common
strains of the virus that are expected to circulate during the upcoming
flu season. It is especially important for individuals who are at higher
risk of severe illness or complications from the flu, such as young
children, older adults, pregnant women, and individuals with underlying
medical conditions.

In terms of the availability and cost, there have been significant
changes for flu vaccines over the past 50 years (CDC, Influenza Historic
Timeline, 2019). In the 1960s, flu vaccines were not widely available
and were primarily given to people considered to be at high risk for
complications. The cost of the vaccine was relatively high and was not
covered by most insurance plans. In the 1970s and 1980s, flu vaccines
became more widely available and were recommended for more people,
including children and healthy adults. The cost of the vaccine began to
decrease as more people received it, and insurance coverage for the
vaccine became more common. In the 1990s and 2000s, the availability of
flu vaccines continued to increase, and the cost of the vaccine became
more affordable for a larger proportion of the U.S. population.

@@

Despite the availability and recommendation of influenza vaccines, many
people in the United States still do not receive them as recommended. In
fact, the CDC reported that only 49.2% of adults aged 18 years and older
received a flu vaccine during the 2020-2021 flu season (CDC, Inequities
in Flu Vaccine Uptake 2022). This low uptake of influenza vaccine
highlights the need for understanding the factors that contribute to
under-coverage of the vaccine in certain populations.

Racial/ethnic minorities have a long history of reduced vaccination
rates relative to white Americans (Khan 2018). Since 2010, flu
vaccination coverage has been consistently lower among Black, Hispanic,
and AI/AN adults. During the 2021--2022 season, flu vaccination coverage
was 54% among White adults, 42% among Black adults, 38% among Hispanic
adults, and 41% among AI/AN adults (CDC, Inequities in Flu Vaccine
Uptake 2022). There are many reasons for these inequities, including
lack of access to health care and insurance, missed opportunities to
vaccinate, and misinformation and distrust.

Racial and ethnic minorities have a longstanding history of lower
vaccination rates compared to white Americans. Since 2010, Black,
Hispanic, and American Indian/Alaska Native adults have consistently had
lower flu vaccination rates than White adults. In the 2021-2022 flu
season, vaccination coverage for White adults was 54%, while it was 42%
for Black adults, 38% for Hispanic adults, and 41% for AI/AN adults
(CDC, Inequities in Flu Vaccine Uptake 2022).  The reasons for these
disparities are multifactorial, and may include limited access to
healthcare and insurance, missed opportunities for vaccination, and
misinformation and mistrust.

In prior studies, lower income levels were associated with decreased
odds of influenza vaccine receipt (Gaskin 2022, Vashist 2022), and
having health insurance coverage has also been shown to be associated
with increased influenza vaccine coverage (Cambou 2021, Vashist 2022).
Influenza vaccination rates were also found to be lower in rural than in
urban areas among adults of all age groups and both genders (Jain 2022).
However, there are still many variables of interest from the NHIS
dataset that have yet to be evaluated in detail. Additionally, the
periods of study from all prior research all end prior to the onset of
the COVID-19 pandemic, which could have altered or heightened
disparities even more.

@@

The data source used for this project is the National Health Interview
Survey (NHIS). It is a harmonized set of data beginning in 1963 with
information on general health status, acute and chronic illness,
functional limitations, access to care, insurance coverage, and health
behaviors for the U.S. population. On average, the survey covers 100,000
persons in 45,000 households each year. It is conducted in-person and by
telephone, and its sample is designed to be representative of the
civilian United States population (NHIS, About).

@@

The raw data for the period of interest directly extracted from the NHIS
IPUMS survey initially had had 792,916 observations. Researchers then
selected their predictors of interest to model the association between
the outcome and variables. Variables assessed are as follows:

@@

@@

```{r, warning=FALSE, message=FALSE}
result_table2 <- readRDS(here("results", "demographics_model.rds"))
result_table2
```

@@

```{r, warning=FALSE}
result_table3 <- readRDS(here("results", "full_model.rds"))
result_table3
```

>>>>

---
title: "Which groups are at highest risk for low uptake of flu vaccine?"
author: Chris Okitondo and Irene Cavros
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../references.bib
csl: ../apa.csl
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(png)

remotes::install_github("paleolimbot/rbbt") #for integration of BibTex Zotero citations
library(rbbt)
```

@@

Influenza, or flu, is an extremely contagious respiratory illness caused by a virus. It can cause a range of symptoms, including fever, cough, sore throat, body aches, headache, chills, and fatigue (@cdc2022c.) It can be particularly dangerous for certain groups, such as older adults, young children, and individuals with underlying health issues, often resulting in severe illness or even death. Although influenza vaccinations are widely accessible, many people in the United States do not receive them as recommended. This study examines influenza vaccination uptake among US adults from 2012-2021. Using a subset of the National Health Interview Survey (NHIS) data set, researchers provide important insights into which groups are at highest risk for under-coverage of flu vaccine in order to better inform targeted public health interventions for improving uptake.

@@

Influenza, commonly known as the flu, is a contagious respiratory illness which comes from influenza viruses. It can cause mild to severe illness, and in some cases, it can lead to hospitalization or even death. Symptoms of the flu include fever, cough, sore throat, body
aches, headache, chills, and fatigue (@cdc2022.) It can spread easily from person to person through droplets that are released infected individuals cough, sneeze, or talk.

Vaccines are a crucial tool in preventing the spread of infectious diseases, including the flu. On an individual level, the best way to protect against the flu is by getting vaccinated each year. At the population level, high vaccine coverage is a powerful strategy for reducing spread of the virus (@cdc2022c.)

The U.S. Centers for Disease Control and Prevention (CDC) recommends that everyone six months of age and older receive a flu vaccine each year, with rare exceptions (@cdc2022c.) Influenza is a constantly evolving virus that mutates rapidly, and uncontrolled spread gives rise to many different strains, so new vaccines are developed each year to protect against the most common strains of the virus that are expected to circulate during the upcoming flu season. It is especially important for individuals who are at higher risk of severe illness or complications from the flu, such as young children, older adults, pregnant women, and individuals with underlying medical conditions.

In terms of the availability and cost, there have been significant changes for flu vaccines over the past 50 years (@InfluenzaHistoricTimeline2022.) In the 1960s, flu vaccines were not widely available and were primarily given to people considered to be at high risk for complications. The cost of the vaccine was relatively high and was not covered by most insurance plans. In the 1970s and 1980s, flu vaccines became more widely available and were recommended for more people, including children and healthy adults. The cost of the vaccine began to decrease as more people received it, and insurance coverage for the vaccine became more common. In the 1990s and 2000s, the availability of flu vaccines continued to increase, and the cost of the vaccine became more affordable for a larger proportion of the U.S. population.

@@

Despite the availability and recommendation of influenza vaccines, many people in the United States still do not receive them as recommended. In fact, the CDC reported that only 49.2% of adults aged 18 years and older received a flu vaccine during the 2020-2021 flu season (@cdc2022b.) This low uptake of influenza vaccine highlights the need for understanding the factors that contribute to under-coverage of the vaccine in certain populations.

Racial/ethnic minorities have a long history of reduced vaccination rates relative to white Americans (@khan_etal2018.) Since 2010, flu vaccination coverage has been consistently lower among Black, Hispanic, and AI/AN adults. During the 2021--2022 season, flu vaccination coverage was 54% among White adults, 42% among Black adults, 38% among Hispanic adults, and 41% among AI/AN adults (@cdc2022b.) There are many reasons for these inequities, including lack of access to health care and insurance, missed opportunities to vaccinate, and misinformation and distrust.

In prior studies, lower income levels were associated with decreased odds of influenza vaccine receipt (@gaskinEffectIncomeDisparities2023; @vashist_etal2022), and having health insurance coverage has also been shown to be associated with increased influenza vaccine coverage (@cambou_etal2021; @vashist_etal2022). Influenza vaccination rates were also found to be lower in rural than in urban areas among adults of all age groups and both genders (@jain_etal2022). However, there are still many variables of interest from the NHIS dataset that have yet to be evaluated in detail. Additionally, the periods of study from all prior research all end prior to the onset of the COVID-19 pandemic, which could have altered or heightened disparities even more.

@@

The data source used for this project is the National Health Interview Survey (NHIS). It is a harmonized set of data beginning in 1963 with information on general health status, acute and chronic illness, functional limitations, access to care, insurance coverage, and health behaviors for the U.S. population. On average, the survey covers 100,000 persons in 45,000 households each year. It is conducted in-person and by telephone, and its sample is designed to be representative of the civilian United States population (@IPUMSNHIS.)

@@

The raw data for the period of interest directly extracted from the NHIS
IPUMS survey initially had 792,916 observations. NHIS Researchers then
selected their predictors of interest to model the association between
the outcome and variables. Variables assessed are as follows:

@@

```{r echo=FALSE, out.width = "50%", fig.align = "center"}
knitr::include_graphics(here("results", "age & flu uptake distribution.png"))
```

@@

```{r, warning=FALSE, message=FALSE, echo=FALSE}
result_table2 <- readRDS(here("results", "demographics_model.rds"))
result_table2
```

@@

```{r, warning=FALSE, echo=FALSE}
result_table3 <- readRDS(here("results", "full_model.rds"))
result_table3
```

### Model Evaluation

#### ROC curve from the full model

```{r echo=FALSE, out.width = "50%", fig.align = "center"}
knitr::include_graphics(here("results", "AUC_full_model.png"))
```

#### ROC curve from the reduced model

```{r echo=FALSE, out.width = "50%",  fig.align = "center"}
knitr::include_graphics(here("results", "AUC_reduced_model.png"))
```
